---
title: "Influence of events"
author: "Mariia Danilenko"
output: 
  html_document:
    keep_md: true
---
##Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

##Questions

Our data analysis will address the following questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

##Data processing.
###Download data with the given link.
```{r, echo=TRUE}
library(R.utils)
library(ggplot2)
library(dplyr)
```

```{r, echo=TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "dataStorm.csv.bz2")
bunzip2("dataStorm.csv.bz2", "dataStorm.csv", remove = FALSE, skip = TRUE)
dataStorm <- read.csv("dataStorm.csv")
head(dataStorm)
```
###Data review.
We are going to use only the following columns from the given data:  
 * EVTYPE - event type  
 * FATALITIES - applies to population health  
 * INJURIES - applies to population health   
* CROPDMG - crop damage  
* CROPDMGEXP - crop damage exponential   
* PROPDMG - property damage  
* PROPDMGEXP - property damage exponential  

Other columns are not applicable to our questions, that's why we will select
only these information for the next steps.
```{r}
selStorm <- select(dataStorm, EVTYPE, FATALITIES, INJURIES, CROPDMG, CROPDMGEXP,PROPDMG, PROPDMGEXP )
head(selStorm)
```


###Convert given damage into human readable numbers.
Two columns (CROPDMGEXP and PROPDMGEXP) should contain encrypt numbers:  
 * K - for thousands;  
 * M - for millions;  
 * H - for hundrites;  
 * B - for billions.  
The initial formula for __Property damage__ is PRPTDMG×PROPDMGEXP.  
The initial formula for __Crop damage__ is CROPDMG×CROPDMGEXP.  

But let's analyze values from these columns:  
```{r}
unique(selStorm$PROPDMGEXP)
unique(selStorm$CROPDMGEXP)
```
Based on the new information construct function switch_vec which helps convert symbols into powers of 10.  
```{r}
switch_vec<- function(x) {
    if (x %in% c('h', 'H'))
        return(2)
    else if (x %in% c('k', 'K'))
        return(3)
    else if (x %in% c('m', 'M'))
        return(6)
    else if (x %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(x)))
        return(as.numeric(x))
    else if (x %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Error")
    }
}
```
Only now we can calculate damage values.  
```{r}
economy <- selStorm
economy$CROPDMGEXP <- sapply(economy$CROPDMGEXP, switch_vec)
economy$PROPDMGEXP <- sapply(economy$PROPDMGEXP, switch_vec)

economy$CROPDMG <- economy$CROPDMG*(10**economy$CROPDMGEXP)
economy$PROPDMG <- economy$PROPDMG*(10**economy$PROPDMGEXP)
```
##Analysis.
### Types of events which are most harmful with respect to population health.

At the begining we have to sum all fatalities and all injuries for each event type.   
For our purpose we don't need to work with events with no fatalities or injuries. Let's filter our data and choose only baneful types of events.  
```{r}
health<- aggregate(cbind(as.numeric(selStorm$FATALITIES),
                         as.numeric(selStorm$INJURIES)), 
                   by=list(selStorm$EVTYPE), FUN="sum" )
health_final<-filter(health, !((V1==0)&(V2==0)))
```

#### Top 10 for fatalities
Arrange our new data to select Top 10 events with the biggest amount of fatalities.  
```{r}
health_Fat<- arrange(health_final, desc(V1))
health_Fat_top <- health_Fat[1:10,]
par(mfcol=c(1,1))
barplot(health_Fat_top$V1, names.arg = health_Fat_top$Group.1, horiz =TRUE, las=1, cex.names = 0.5, main="Events with Top 10 fatalities")
```

#### Top 10 for injuries
Arrange our new data to select Top 10 events with the biggest amount of injuries.  
```{r}
health_Inj<- arrange(health_final, desc(V2))
health_Inj_top <- health_Inj[1:10,]
par(mfcol=c(1,1))
barplot(health_Inj_top$V2, names.arg = health_Inj_top$Group.1, horiz =TRUE, las=1, cex.names = 0.5, main="Events with Top 10 injuries")
```
  
### Types of events which have the greatest economic consequences  
Aggregate two types of damage by event type.  
```{r}
economy_final<- aggregate(cbind(economy$CROPDMG,economy$PROPDMG), 
                   by=list(economy$EVTYPE), FUN="sum" )
```
At this moment we can select Top 5 events for each type of damage:    
```{r}
par(mfcol=c(2,1))

economy_final <- arrange(economy_final, desc(V1))
economy_crop=economy_final[1:5,]
barplot(economy_crop$V1/economy_crop$V1[1], names.arg = economy_crop$Group.1, horiz =TRUE, las=1, cex.names = 0.5,col="lightgreen", main="Events with Top 5 Crop damage", xlab="Numbers are given in 1.4e+10 US dollars")

economy_final <- arrange(economy_final, desc(V2))
economy_prop=economy_final[1:5,]
barplot((economy_prop$V2 / economy_prop$V2[1]), names.arg = economy_prop$Group.1, horiz =TRUE, las=1, cex.names = 0.5,col="lightblue", main="Events with Top 5 Property damage",xlab="Numbers are given in 6.8e+13 US dollars")
```

##Results.  
###Types of events which are most harmful with respect to population health  
The Top 10 for fatalities is:  
```{r}
names(health_Fat_top)<-c("EvType", "Fatalities", "Injuries")
health_Fat_top[1:10,]
```
The Top 10 for injuries is:  
```{r} 
names(health_Inj_top)<-c("EvType", "Fatalities", "Injuries")
health_Inj_top[1:10,]
```
We can't compare fatality and injury that's why we won't analyse the sum.  
Also, we can recognize that Tornado is the most influential event is the US.   Notice, that we calculated the sums of deaths and sufferers not the means.    
### Types of events which have the greatest economic consequences  
The Top 5 for Crop damage is:
```{r}
names(economy_crop)<-c("EvType", "CropDMG", "PropDMG")
economy_crop[1:5,]
```
The Top 5 for Property damage is:
```{r}
names(economy_prop)<-c("EvType", "CropDMG", "PropDMG")
economy_prop[1:5,]
```

As we can see Crop damage is much less than Property damage and doesn't impact to Total damage (It can be easily explained).    
Based on the given statistics the flash flood have the greatest economic consequences.  
